---
title: "Getting Started with rapR: A Guide for Music Psychology Researchers"
author: "Josh Gonzales"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rapR: A Guide for Music Psychology Researchers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction: Why rapR?

As a music psychology researcher, you might want to analyze song lyrics, track artist evolution, or study music trends. **rapR** makes this possible without needing to be a programmer. This guide walks you through everything step-by-step.

## What You'll Learn

By the end of this guide, you'll be able to:

1. Set up rapR and get your API credentials
2. Search for artists and songs
3. Collect data about an artist's entire discography
4. Export your data to Excel or CSV for analysis
5. Create basic visualizations of music data

**No prior programming experience needed!** Just copy and paste the code examples.

## Step 1: Installation

First, install rapR. Copy and paste this into your R console:

```{r install, eval=FALSE}
# Install the package that lets us install from GitHub
install.packages("remotes")

# Install rapR
remotes::install_github("phdemotions/rapR")

# Load the package
library(rapR)
```

## Step 2: Getting Your Genius API Token

To use rapR, you need a free API token from Genius. Don't worry - it's simple!

### How to Get Your Token:

1. Go to https://genius.com/api-clients
2. Sign in (or create a free account)
3. Click "New API Client"
4. Fill in:
   - **App Name**: "My Research Project" (or anything you like)
   - **App Website URL**: "https://yourwebsite.com" (can be anything)
5. Click "Save"
6. Click "Generate Access Token"
7. Copy the long string of letters and numbers - this is your token!

### Setting Your Token in R:

```{r set-token, eval=FALSE}
# Replace YOUR_TOKEN_HERE with the token you copied
set_genius_token("YOUR_TOKEN_HERE")
```

**Pro Tip**: To avoid entering your token every time, add it to your `.Renviron` file. Run this once:

```{r save-token, eval=FALSE}
# This opens a file where you can save your token permanently
usethis::edit_r_environ()

# In the file that opens, add this line (with YOUR actual token):
# GENIUS_API_TOKEN=your_token_here

# Save the file and restart R
```

## Step 3: Your First Search - Finding an Artist

Let's search for an artist. We'll use Kendrick Lamar as an example:

```{r search-artist, eval=FALSE}
# Search for Kendrick Lamar
artist_data <- genius_get_artist_details_df("Kendrick Lamar")

# View the data
print(artist_data)
```

You'll see information like:
- Artist ID (you'll need this later!)
- Artist name
- Genius URL
- Verification status
- IQ score (Genius's measure of community contributions)

**If multiple artists appear**: The function will ask you to choose the correct one - just type the number and press Enter!

## Step 4: Finding a Specific Song

Want data about a specific song?

```{r search-song, eval=FALSE}
# Search for the song "HUMBLE."
song_data <- genius_get_song_details_df("HUMBLE")

# View the data
print(song_data)
```

You'll get:
- Song title
- Artist name
- Release date
- URL to the Genius page
- Page views (popularity metric!)
- Lyrics state (whether lyrics are complete)

## Research Example: Analyzing an Artist's Career

Here's a complete research workflow. Let's say you want to study how an artist's popularity has changed over time.

### Collecting All Songs by an Artist

```{r get-all-songs, eval=FALSE}
# First, get the artist's ID from your previous search
# For this example, let's say Kendrick Lamar's ID is 1421
artist_id <- "1421"  # Replace with the actual ID from your search

# Get ALL songs by this artist
all_songs <- genius_get_all_songs_from_artist(artist_id)

# View the first few rows
head(all_songs)

# How many songs did we get?
nrow(all_songs)
```

**Note**: This might take a few minutes for artists with many songs. The function automatically pauses between requests to be respectful to Genius's servers.

### Saving Your Data

Now save this data so you can analyze it in Excel, SPSS, or other tools:

```{r save-data, eval=FALSE}
# Save as Excel file
library(writexl)
write_xlsx(all_songs, "kendrick_songs.xlsx")

# OR save as CSV
write.csv(all_songs, "kendrick_songs.csv", row.names = FALSE)
```

### Basic Analysis in R

Even without advanced R skills, you can do simple analyses:

```{r basic-analysis, eval=FALSE}
# Load dplyr for easy data manipulation
library(dplyr)

# Count how many songs per album
songs_per_album <- all_songs %>%
  group_by(full_title) %>%
  summarise(count = n())

print(songs_per_album)

# Get summary statistics
summary(all_songs)
```

## Research Example: Comparing Multiple Artists

Want to compare different artists? Here's how:

```{r compare-artists, eval=FALSE}
# Get data for multiple artists
artist1 <- genius_get_artist_details_df("Kendrick Lamar")
artist2 <- genius_get_artist_details_df("J. Cole")
artist3 <- genius_get_artist_details_df("Drake")

# Combine into one dataset
library(dplyr)
comparison <- bind_rows(
  artist1 %>% mutate(artist_searched = "Kendrick Lamar"),
  artist2 %>% mutate(artist_searched = "J. Cole"),
  artist3 %>% mutate(artist_searched = "Drake")
)

# Save for analysis
write_xlsx(comparison, "artist_comparison.xlsx")
```

## Research Example: Studying Song Lyrics and Annotations

Genius is famous for its crowdsourced lyric annotations. Here's how to access them:

```{r annotations, eval=FALSE}
# Search for a song and get its details
song <- genius_get_song_details_df("Alright")

# Get the song ID from the results
song_id <- song$id

# Get all annotations (interpretations) for this song
annotations <- genius_get_referents(song_id = song_id)

# View the annotations
print(annotations)
```

This gives you access to:
- What parts of lyrics are annotated
- Community interpretations
- Number of annotations per song (measure of cultural impact!)

## Creating Simple Visualizations

Even basic visualizations can reveal interesting patterns:

```{r visualizations, eval=FALSE}
# Load ggplot2 for visualization
library(ggplot2)

# Visualize page views across songs (if you have that data)
# Assuming you have a dataset with song titles and page views
ggplot(song_data, aes(x = release_date, y = page_views)) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "Song Popularity Over Time",
    x = "Release Date",
    y = "Page Views"
  ) +
  theme_minimal()
```

## Research Applications

Here are some research questions you can answer with rapR:

### Music Psychology Research:
- How do emotional themes in lyrics change across an artist's career?
- What makes some songs more culturally significant (more annotations)?
- How does audience engagement (page views) relate to critical acclaim?

### Cultural Studies:
- How do different genres use language differently?
- What social issues are referenced in popular music?
- How has the diversity of voices in music changed over time?

### Computational Analysis:
- Build datasets of thousands of songs for machine learning
- Track trends in music over decades
- Compare artists across genres

## Troubleshooting Common Issues

### "Request failed with status: 401"
Your API token is incorrect or expired. Get a new one from Genius and use `set_genius_token()` again.

### "Song not found in search results"
Try a more specific search. For example:
- Instead of "Humble", try "HUMBLE Kendrick Lamar"
- Check the spelling
- Try searching on Genius.com first to see how the song is listed

### Too many results to choose from
The function will show you options. Just type the number of the correct one!

### Data looks complicated
Start small! Extract just the columns you need:

```{r simplify, eval=FALSE}
# Select just the columns you want
library(dplyr)
simple_data <- all_songs %>%
  select(title, artist, release_date, url)
```

## Next Steps

### Learn More R:
- [R for Data Science](https://r4ds.had.co.nz/) - Free online book
- [RStudio Education](https://education.rstudio.com/)

### Explore Other rapR Functions:
- `genius_get_search_results()` - General search
- `genius_get_web_page()` - Get web page data
- `genius_get_annotation()` - Get specific annotations

### Join the Community:
- Report bugs: https://github.com/phdemotions/rapR/issues
- Ask questions: Open a GitHub issue with the "question" label
- Share your research: We'd love to hear how you use rapR!

## Example Research Workflow: Complete Project

Here's a complete example from start to finish:

```{r complete-example, eval=FALSE}
# 1. Load packages
library(rapR)
library(dplyr)
library(writexl)

# 2. Set your token
set_genius_token("YOUR_TOKEN_HERE")

# 3. Research question: How has Drake's discography grown over time?

# Get Drake's information
drake <- genius_get_artist_details_df("Drake")
drake_id <- drake$id

# Get all his songs
drake_songs <- genius_get_all_songs_from_artist(drake_id)

# 4. Get details for top 10 songs
top_songs <- drake_songs %>%
  head(10)

# 5. Get detailed information for each song
song_details <- data.frame()

for (i in 1:nrow(top_songs)) {
  song_name <- top_songs$title[i]

  # Get details
  details <- genius_get_song_details_df(song_name)

  # Add to our dataset
  song_details <- bind_rows(song_details, details)

  # Pause between requests
  Sys.sleep(2)
}

# 6. Save your data
write_xlsx(list(
  "Artist Info" = drake,
  "All Songs" = drake_songs,
  "Top Songs Details" = song_details
), "drake_research_data.xlsx")

# 7. You now have data ready for analysis in Excel, SPSS, or other tools!
```

## Citation

If you use rapR in your research, please cite it:

```{r citation, eval=FALSE}
citation("rapR")
```

This helps us show the impact of the package and keep it maintained!

## Conclusion

You don't need to be a programmer to use rapR for your research. Start with the examples in this guide, modify them for your needs, and don't be afraid to experiment. The R community is friendly and helpful!

**Happy researching!**
